---
title: "EMC MST Ballistic Accumulation Model"
output: html_notebook
---

This note book will recreate the EMC model design framework with the MST task in mind
```{r}
library(EMC2)
#.rs.restartR()
```

Step 1: Loading and examining data
```{r}
data <- read.csv("C:\\Users\\thaug\\Downloads\\mst_all_data_model.csv") # String to read csv from

data <- data[!(data$trial_index %in% c(0, 1, 130)),]

tail(data)
```

We then want to create an appropriate datafrmae to feed into the model, whcih means we will generate group tags as they have not been made available in the current data

```{r}
group_names <- c("yhc", "ahc", "cl") # Creating a list of our options

# We then want to give every subject a category they belong in
subjects <- unique(data$subject)
set.seed(42)
group_assignments <- sample(group_names, length(subjects), replace=TRUE)
names(group_assignments) <- subjects

# Mapping the assignment to the dataframe
assignments <- group_assignments[data$subject]
print(assignments)
  
# I then start to think about what parts we need to represent in our MST data
# for clarity we create a dataframe containing all parts of the data we intend to use
# with the current model - when thinking about changing the model we can go back up to 
# the original imported frame and add them from there

data_model1 <- data.frame(
  subject = data$subject, # the ID of the participant 
  G = assignments, # the clinical group the participant is in
  S = data$correct_response, # the correct response to the stimulus presented
  lR = data$resp, # the response provided by the participant
  rt = data$rt/1000 # the response time of the participants response changed to seconds
)


tail(data_model1)
```

NB! lRemember to make it siongle for the single person model



Step 2: Making the specifications for the model

For race models we need to specify a function to connect the the participant responses with the correct response

```{r}
# Giving the match function for the ballistic model for MST
#matchfun <- function(data) data$S == data$lR # Creating a link between response and correct


# But because there is 3 potential responses we need to map them unto the model so that there can be 3 accumulators with another match function to account for this
responses = c("new", "old", "similar")
# Taking the data from our dataframe
matchfun <- function(data_model1){
  as.character(data_model1$lR) # lReturning the object response to code it in as the correct 

  
plot_density(data_model1, factors = c("rt"), functions = list(correct = matchfun), defective_factor = "correct")
}
head(data_model1)
```

To make it into a one subject analysis we extract the first subject from the dataset 
```{r}
data_single <- data_model1[data_model1$subject == "MST-281",] # Selecting the first subject
#data_single$subjects <- factor(data_single$subjects)
tail(data_single)
```

From what I can gather for this model it is useful to specify and average and difference matrix - this is factors, but as they are specified in the dataframe that might not be necessary from waht I can tell

```{r}
ADmat <- cbind(
  d1 = c(1, -1/2, -1/2),
  d2 = c(0, sqrt(3)/2, -sqrt(3)/2)
)
rownames(ADmat) <- c("n", "o", "s") # For new, old, and similar stimulus
ADmat
```


We then sepcify the formula for the model
```{r}
# Creating a formula for MST we use the following shorthands
# G is the participant group
# S = variable stating correct response
# a ~ E - finding the intercept for the condition
# t0 ~ 1 estimates a single parameter each for all of the cells for non-decision time
# Z - response bias
# sv between trial evidence-accumulation rate
# st0 - non-decision time
# SZ - starting points
# B ~ G2 - thresholds for the accumulators which are allowed to vary according
# to which groups aka accumulators 
# A represents teh start point variability ranging from 0 to A

# Specifying the design for the model
G3 <- function(data) factor(data_single$G!="group",labels=c("yhc","ahc", "cl"))
design_MST <- design(data = data_single,model=LBA,matchfun=matchfun,
                        formula=list(v~lR+G3,sv~1,B~lR+G3,A~1,t0~1),
                        contrasts=list(lM = ADmat),constants=c(sv=log(1)), 
                        functions=list(G3=G3))
```

We then need to set the priors for this model
```{r}
prior_mean_MST <- c(v_Sleft=-2, v_Sright=2, a=log(1), a_Eneutral=log(1.5),
                a_Eaccuracy=log(2), t0 = log(.2), Z=qnorm(.5), sv=log(.5))

prior_mean_MST <- c(v = 1, v_lR = 1, )

prior_mean_MST <- c(
  # Setting drift rates for combination of response and accumulator
  "v_new:yhc" = 2,
  "v_old:yhc" = 2,
  "v_similar:yhc" = 2,
  
  "v_new:ahc" = 2,
  "v_old:ahc" = 2,
  "v_similar:ahc" = 2,
  
  "v_new:cl" = 2,
  "samples/LBABvE2_nollR.lRData:cl" = 2,
  "v_similar:cl" = 2,
  
  # Settting starting point variability (sv)
  "sv_new" = log(1),
  "sv_old" = log(1),
  "sv_similar" = log(1),
  
  # Boundary separation
  "B_yhc" = log(1.5),
  "B_ahc" = log(1.5),
  "B_cl" log(1.5),
  
  # Setting non-decison time and start point variability
  "A" = log(0.2),
  "t0" = log(0.35)
)

prior_sd_MST <- rep(0.5, length(pmean))  # simply giving the same sd to all parameters
# to keep it simple in the beginning

# Construct prior
prior_MST <- prior(design_MST,pmean = prior_mean_MST, psd = prior_sd_MST,
                     type = "single")


```

We then plot the mapped priors
```{r}
plot(prior_MST) # Add map = F to look at them unmapped
```


Part 3: Creating and fitting model object

```{r}
LBA_MST <- make_emc(data, design_MST, prior = prior_MST)
LBA_MST<- fit(LBA_MST)
check(LBA_MST) # Checking how the model converges as well as efficiency (still escapes me tho)
```



