---
title: "Summer School Prep"
output: html_notebook
---

This notebook will do teh prep-work necessary ot try to be ready for 2025 model based
neuroscience summer school

First we load the package

```{r}
#install.packages("EMC2")
library(EMC2)
```
The constraints of this package is as follows:
- Data needs to be in a data.frame format
- Column containing participant identifiers
- Column containing responses (titled R)
- Column containing response time in seconds (rt)

I then downloaded the LVABvE, DDMavE and DDMae into my environment for data
(from "osf.io/7euya/" website)

First we need to load the data with the example dataset in the package
and examine it
```{r}
dat <- forstmann
head(dat)

```

In the data where correctness is not added automatically we need to create
it 
```{r}
plot_density(dat, factors = c("E"), functions = list(correct = correct_fun), defective_factor = "correct")


```



To get a visual impression of the data we have loaded
```{r}
plot_density(dat, factors = c("S", "E"), layout = c(2,3))
```

To make it into a one subject analysis we extract the first subject from the dataset 
```{r}
dat_single <- dat[dat$subjects == levels(dat$subjects)[1],] # Selecting the first subject
dat_single$subjects <- factor(dat_single$subjects)
```


We then sepcify the formula for the model

```{r}
formula <- list(v~0+S, a~E, t0~1, s~1, Z~1, sv~1, st0~1, SZ~1)
# S = variable stating correct response
# a ~ E - finding the intercept for the condition
# t0 ~ 1 estimates a single parameter each for all of the cells for non-decision time
# Z - response bias
# sv between trial evidence-accumulation rate
# st0 - non-decision time
# SZ - starting points


# Constants are specified at the rate of which sampling takes place

# We then fit the formula to the data - that is wrong
# We instead specify the model design
design_DDMaE <- design(data = dat_single, model=DDM,
                       constants = c(s=log(1)), # Variability set to constant because there is only one participant currently
                       formula = formula)



?design
```

To see how the function maps the variables there is an own function for doing that

```{r}
mapped_pars(design_DDMaE)
#mapped_pars(design_MST)
```

We then move along to the Baeysian specification part of the model where we 
- Apparently when there is a single subject model the package assumes normal priors

We then set the the priors which the model uses

```{r}
# Set prior mean
prior_mean <- c(v_Sleft=-2, v_Sright=2, a=log(1), a_Eneutral=log(1.5),
                a_Eaccuracy=log(2), t0 = log(.2), Z=qnorm(.5), sv=log(.5))

# Set prior standard deviation
prior_sd <- c(v_Sleft=1,v_Sright=1,a=.3,a_Eneutral=.3,a_Eaccuracy=.3,
           t0=.4,Z=1,sv=.4,SZ=1)

# Construct prior
prior_DDMaE <- prior(design_DDMaE,pmean = prior_mean, psd = prior_sd,
                     type = "single")
```


We then display the priors for visual inspection

```{r}
plot(prior_DDMaE, layout = c(2,4))
```
When using different models there is a compare function which can be used to look at them in the package
- Always a good idea to display beforehand tho


What we need to do to get the trials sorted correctly is to parse them into correct and faulty trials (something about that in one of the lectures but not sure the degree of which it applies)


To make it into an object (I think) with the make_emc function
```{r}
DDMaE <- make_emc(dat_single,design_DDMaE, prior=prior_DDMaE)
```
We then fit the model object to our data - which might take a jiffy

```{r}
DDMaE <- fit(DDMaE)
```
After having run this there are some tests which could be useful to do
Including:
```{r}
?check
check(DDMaE)
```

Summary:
```{r}
summary(DDMaE)
```
And comparison of priors and posteriors

```{r}
?plot_pars
plot_pars(DDMaE)
```

After having done this we then want to see how the model fits the data

In which we first use the predict function to use the model to predict our data
and plot the different cdfs for comparison
```{r}
?predict
pp_DDMaE <- predict(DDMaE)
plot_cdf(DDMaE, pp_DDMaE)
```

