---
title: "Introduction to EMC2"
author: "Niek Stevenson"
description: A brief introduction to fitting evidence accumulation models with EMC2
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(comment = "")
```

In this introduction we will walk you through the basic steps of fitting
evidence accumulation models with EMC2. The document is meant as a
software introduction to the package, and is not meant to be
comprehensive in terms of the theoretical background.

Before proceeding with this R script, we strongly recommend reading the
EMC2 paper which provides a more comprehensive overview of the inner
workings and workflow of the package.

To start with let's load in EMC2 and set a seed for reproducibility. 

```{r setup}
rm(list=ls())
# install.packages("EMC2")
library(EMC2)
RNGkind("L'Ecuyer-CMRG")
set.seed(123456)
```

# Data preparation

For this tutorial, we will analyze data from a seminal experiment by
Forstmann et al. (2008). Their study investigated how striatum and
pre-SMA facilitate decision-making under time pressure, published in the
Proceedings of the National Academy of Sciences
(<https://doi.org/10.1073/pnas.0805903105>).

The dataset is included in the EMC2 package as the "forstmann"
dataframe. Let's create a working copy:

```{r load_data}
dat <- forstmann
```

## Data Format

EMC2 has specific requirements for column names: - "rt" for response
times (in seconds) - "R" for responses - "subjects" for participant
identifiers

Note that all names are case-sensitive.

Let's examine our data:

```{r examine_data}
head(dat)
```

In this dataset, the response factor (R) and stimulus factor (S) share
the same levels: "left" and "right". This makes it straightforward to
score correct responses by checking if S equals R. The main experimental
manipulation involves cues presented before each trial, which instruct
participants to emphasize either speed, accuracy, or remain neutral.
These instructions are captured in the E (emphasis) factor with three
levels: "speed", "neutral", and "accuracy".

The data reveals clear speed-accuracy trade-offs across conditions:

-   Under speed instructions: Participants respond faster but make more
    errors

-   Under accuracy instructions: Participants respond slower but achieve
    higher accuracy

-   Under neutral instructions: Performance falls between the speed and
    accuracy conditions, but closer to the accuracy pattern

```{r plot_density, fig.width = 6, fig.height = 4}
plot_density(dat, factors = c("S", "E"), layout = c(2,3))
```

Note: This data was previously truncated at 0.25s and 1.5s to remove
outliers.

# Diffusion Decision Model (DDM)

## Single Subject Model

We'll begin by demonstrating how to fit a model to a single subject
before moving on to hierarchical modeling. Let's extract data for only
the first subject:

```{r single_subject_data}
dat_single <- dat[dat$subjects == levels(dat$subjects)[1],]
dat_single$subjects <- factor(dat_single$subjects)
```

We'll now specify a DDM model using the Forstmann data. We will start
with the most intuitive parameterization, where only the threshold
parameter varies with emphasis condition (E).

Our model includes several key parameters:

-   Drift rates (v): One each for leftward and rightward moving stimuli

-   Threshold (a): Varies with the emphasis factor

-   Non-decision time (t0): Fixed across conditions

-   Starting point (z): Fixed across conditions

-   Drift variability (sv): Fixed across conditions

For drift rates, we use dummy coding (v\~0+S) to estimate separate rates
for left and right stimuli. The first level (left) maps to the bottom
boundary and the second (right) to the top, so negative rates favor left
responses and positive rates favor right responses.

Following the selective influence assumption, we allow only the
threshold parameter (a) to vary with the speed-accuracy emphasis
manipulation (a\~E). This parameter, also known as "response caution" in
DDM literature, changes based on the emphasis condition:

-   Speed condition serves as the baseline level (intercept)

-   Two additional parameters capture threshold differences:

    -   Neutral vs. speed

    -   Accuracy vs. speed

All other model parameters remain constant across conditions, with one
value estimated per parameter. Since EMC2 assumes that parameters are
sampled on the real line, we use the log transformation for a and t0 to
enforce that they are positive, and the probit transformation for Z,
which is estimated in a proportional sense (i.e., Z = z/a).

To ensure model identifiability, we fix the diffusion coefficient (s) to
log(1) using the constants argument. We specify log(1) since s is also
estimated on the log scale as it must be positive. On the natural scale
we have thus fixed s to 1. To see all the parameter transformations, see
`?DDM`.

Any parameters not explicitly specified in the formula receive their
default values, with message telling you what they are.

### Design Specification

The `design` function is arguably the most important function in EMC2.
Here we specify the model type (DDM) and a data object. When the data is
properly formatted, the function automatically extracts important design
information that can also be specified manually. This includes:

-   Response levels (`Rlevels`): The possible response options, in this
    case `c("left", "right")`
-   Factors (`factors`): A list defining the experimental factors and
    their levels, such as:
    -   Stimulus (S): `c("left", "right")`
    -   Emphasis (E): `c("speed", "neutral", "accuracy")`

If the data contains numerical columns (besides rt) design will included
them as a covariate. We recommend not including any unnecessary columns
in your data as they can slow down model estimation.

When we run `design`, it prints the names of parameters to be sampled (8
in this case) and their corresponding design matrices. To reflect the
selective influence assumption we call this "DDMaE":

```{r design_DDMaE, comment = ""}
design_DDMaE <- design(data = dat_single,model=DDM,
                       formula =list(v~0+S,a~E, t0~1, s~1, Z~1, sv~1),
                       constants=c(s=log(1)))
```

The print and summary functions help inspect the design at different
levels of detail:

```{r print_design_DDMaE, eval = FALSE}
print(design_DDMaE)
```

For a more detailed output showing all design matrices:

```{r summary_design_DDMaE}
summary(design_DDMaE)
```

Understanding how your parameterization maps to the experimental design
can be challenging. The `mapped_pars` function helps build this
intuition:

```{r mapped_pars_DDMaE}
mapped_pars(design_DDMaE)
```

This will print a description of the transformations and mappings of the
parameters back to the experimental design for all parameters that vary
as a function of the experimental factors.

We can also make an illustration of the proposed design. In this case we
need to construct a parameter vector ("p_vector"). To do so, we first
construct a vector of example parameters, on the scales they are sampled
(thus keeping the parameter transformations in mind).

Here we'll use some reasonable values that will later also be used to
specify a prior.

```{r p_vector_DDMaE}
p_vector=c(v_Sleft=-2,v_Sright=2,a=log(1),a_Eneutral=log(1.5),a_Eaccuracy=log(2),
           t0=log(.2),Z=qnorm(.5),sv=log(.5))
```

Now we can plot the design. This plot is for visual purposes, although
it matches the actual proposed accumulation process of the DDM as
closely as possible, some parts are altered for clarity. You might need
to adjust the plot size to fit your display:

```{r plot_DDMaE}
plot(design_DDMaE, p_vector, factors = list(v = "S", a = "E"), plot_factor = "S")
```

Side note: you can also create an empty parameter vector as follows
(doMap=FALSE removes the design matrices to simplify the output):

```{r sampled_pars_DDMaE}
sampled_pars(design_DDMaE,doMap=FALSE)
```

### Prior Specification

Single subject models assume normal priors on the parameters, requiring
us to specify means and standard deviations for each of the 8 model
parameters.

Note that priors are specified on the sampled scale, which requires us
to keep the parameter transformations in mind when specifying the prior.
As an example, we use an informed prior based on:

1.  Expecting fairly accurate responding (negative rate for left,
    positive for right)
2.  Typical DDM drift rate values around 2
3.  Response caution decreasing from accuracy to neutral to speed
    instruction

For the prior on parameter means, we use the p_vector values defined
above. Again, as an example, we found the following standard deviations
to be reasonable:

```{r prior_sd_DDMaE}
prior_sd <- c(v_Sleft=1,v_Sright=1,a=.3,a_Eneutral=.3,a_Eaccuracy=.3,
           t0=.4,Z=1,sv=.4,SZ=1)

prior_DDMaE <- prior(design_DDMaE,pmean = p_vector, psd = prior_sd, 
                     type = "single")
```

We set type is "single" because we are fitting a single subject model
and we need to create a matching prior. Just as with the design, the
print and summary functions can be used to inspect the prior with
different levels of detail.

```{r print_prior_DDMaE, eval = FALSE}
print(prior_DDMaE)
```

Or for a more detailed output:

```{r summary_prior_DDMaE}
summary(prior_DDMaE)
```

To check if our prior is reasonable, we can visualize it using the plot
function. The default plotting behavior maps parameters back to their
natural scales.

Visualizing this mapping is important because parameter transformations
can have counterintuitive effects. For example, when using the probit
transformation, a normal prior with mean=0 and SD=1 actually creates a
uniform prior on the mapped scale.

```{r plot_prior_DDMaE}
plot(prior_DDMaE)
```

Note that there is slight un-evenness in the plots as they are obtained
by simulation, which is why these functions can take some time to run.

We can also plot the prior on the unbounded scale, using the map=FALSE
option. This will show the normal distributions on the sampled scale.

```{r plot_prior_DDMaE_unbounded}
plot(prior_DDMaE, map = FALSE)
```

Similarly, we can also inspect quantiles of the prior using the
`credint` function.

```{r credint_DDMaE}
credint(prior_DDMaE)
```

Lastly we can again inspect the proposed mapping using the mapped_pars
function, which now uses the prior means as inputs:

```{r mapped_pars_prior_DDMaE}
mapped_pars(prior_DDMaE)
```

This returns a matrix with columns for every parameter type. For
example, the threshold parameter 'a' is mapped as:

-   Speed condition: exp(log(1)) = 1

-   Neutral condition: exp(log(1) + log(1.5)) = 1.5

-   Accuracy condition: exp(log(1) + log(2)) = 2

The output also includes additional parameters that aren't directly
specified but can be useful for interpretation. Note, for example, the
difference between relative bias (Z) (which is proportional to the
response boundary a) and absolute bias (z).

Our prior looks reasonable, so we can now fit the model.

### Model Estimation

First, we combine the data, design and prior in an emc object. By
default `make_emc` creates an object with three chains. It assumes that
response time is recorded with a resolution of 0.02 seconds (assuming a
50Hz refresh rate of a monitor and hence in most experimental setups the
maximum resolution for recording response times to visual stimuli).

This specification compresses the data to speed up sampling by:

1.  Keeping only rows with unique combinations of parameters, responses
    and rt values
2.  Computing log-likelihood calculations only for these unique values
3.  Replicating and summing the log-likelihoods appropriately

Since we're only fitting a single subject model, we again specify the
single type.

```{r make_emc_DDMaE}
DDMaE <- make_emc(dat_single, design_DDMaE, prior = prior_DDMaE, type = "single")
```

Next we run the parameter estimation (sampling). By default 3 x 1000
samples will be obtained after burn in and adaptation (but possibly more
if this is not enough to satisfy default convergence criteria). By
default EMC2 parallelizes across the (in this case 3) chains. This can
be turned off by setting cores_for_chains = 1.

The fileName argument saves intermediate progress (as an object called
"emc") to a file, which is useful for longer runs in case they need to
be restarted. Sampling can be slow so you just load the results instead
of running the next chunk (this is more so for the hierarchical modeling
we will look at next).

```{r fit_DDMaE, eval=FALSE}
DDMaE <- fit(DDMaE,  fileName = 'tmp.RData')
save(DDMaE, file = "DDMaE.RData")
```

```{r load_DDMaE, eval=TRUE, echo=FALSE}
load("DDMaE.RData")
```

Notice how sampling sped up towards the end of the sampling phase?
That's because the number of particles is adaptively scaled down as the
proposals get more concentrated around the posterior mode.

The print function gives a quick overview of the estimated model:

```{r print_DDMaE}
print(DDMaE)
```

The check function serves two purposes: 1. Verifies that sampling has
converged 2. Quantifies sampling efficiency through the "effective"
number of samples (ESS)

ESS represents the number of samples after discounting auto-correlation.
When performing posterior inference (e.g., estimating 95% credible
intervals), it's important to have sufficient effective samples for
stable estimates.

```{r check_DDMaE, fig.width = 4, fig.height = 4}
check(DDMaE)
```


The check function also displays plots of the worst converging chains
for each parameter type (in this case only the `alpha` type, since we
fit a single subject model), as measured by the Gelman-Rubin statistic
(Rhat/psrf). This makes it easy to visually identify any convergence
issues in the sampling.

For a more comprehensive analysis of the MCMC chains, we can use the
summary function. This provides detailed statistics for each parameter,
including the Rhat, ESS, and key quantiles of the posterior
distribution.

```{r summary_DDMaE}
summary(DDMaE)
```

We can also look at the full set of MCMC chains which show well
converged the chains are.

```{r plot_DDMaE_chains}
plot(DDMaE)
```

### Model Inference

The plot_pars function can be used to plot posterior densities for each
parameter (black) superimposed on priors (red) to check how much the
priors have been "updated" by the data. Ideally, we want priors to be
"dominated" by the data, meaning the red line is relatively flat while
the black line is sharply peaked.

The dominance is quantified by a contraction statistic: one minus the
ratio of posterior to prior variance. Values close to 1 indicate that
the prior has been effectively washed out by the data.

```{r plot_pars_DDMaE, fig.height = 6, fig.width = 6}
plot_pars(DDMaE)
```


Note that the prior on `sv` wasn't nearly as washed out as the other
parameters, indicating that the data didn't provide much information
about this parameter.

We can also plot the parameter posteriors as they are mapped back to the
original design:

```{r plot_pars_DDMaE_map, fig.height = 6, fig.width = 6}
plot_pars(DDMaE,map=TRUE)
```

We can assess how well the model fits the data visually using posterior
predictives, which are calculated by:

1.  Sampling parameter sets from the posterior
2.  Using these parameters to simulate new data

This can be slow, so using parallel processing through the n_cores
argument is a good idea.

```{r predict_DDMaE}
pp_DDMaE <- predict(DDMaE, n_cores = 12)
```

We can quantify misfit using defective cumulative distribution function
plots. Here we add a function to the data that creates a new column
"correct":

```{r plot_cdf_DDMaE, fig.height = 4, fig.width = 7}
acc_fun <- function(data) return(data$S == data$R)
plot_cdf(DDMaE, pp_DDMaE, factors = "E", functions = c(correct = acc_fun), 
         layout = c(1,3),defective_factor = "correct", 
         legendpos = c("bottomright", "right"))
```

In these plots:

-   Green line: Median of predicted data

-   Light green shading: 95% credible intervals of the posterior
    predictives

-   Black line: Observed data

-   X-axis: Response time in seconds

-   Y-axis: Cumulative defective density (probability of response by
    that time)

Similarly, we can plot the defective density of the posterior
predictives over the real data:

```{r plot_density_DDMaE, fig.height = 4, fig.width = 7}
plot_density(DDMaE, pp_DDMaE, factors = "E", functions = c(correct = acc_fun),
            layout = c(1,3),defective_factor = "correct",
            legendpos = c("topright", "right"))
```

Note that our model does not seem to fit the data well. Most notably,
response times are overestimated and number of error responses is
underestimated

We can also include prior predictives in `plot_cdf` or `plot_density`.
To do this: we set `to_plot = c('real', 'posterior', 'prior')`.
Alternatively, we can create and save prior predictives using
`predict(prior_DDMaE)`. By default, plot limits are based on real data
and posterior predictives. Include prior predictives in limits using the
use_lim parameter

```{r plot_cdf_DDMaE_prior, fig.height = 4, fig.width = 7}
plot_cdf(DDMaE, pp_DDMaE, factors = "E", functions = c(correct = acc_fun), 
         layout = c(1,3),defective_factor = "correct", 
         legendpos = c("topright", "topleft"), 
         to_plot = c('real', 'posterior','prior'))

# or: priorp_DDMaE <- predict(prior_DDMaE)
# plot_cdf(DDMaE, pp_DDMaE, priorp_DDMaE, ...)
```

Obviously our prior predictions are quite broad.

## Hierarchical Model

We will now move on to hierarchical estimation of the DDM. Given the
poor fit (albeit on just one subject), let's break the conventional
selective influence assumption and allow response emphasis (E) to also
influence drift rates.

Instead of using different drift rates for left and right stimuli, for
each E level we will reparameterize the drift rates into:

1.  A drift bias: The average rate across stimuli
2.  A drift rate difference: The difference between stimuli

This can be achieved by specifying a contrast matrix for the S factor
that takes the difference between right and left stimuli:

```{r Smat_DDMaE}
Smat <- matrix(c(-1,1), nrow = 2,dimnames=list(NULL,"dif"))
```

To see how this works, we alter the previous design with the formula
v\~S, supplying Smat in the contrasts argument:

```{r new_design_DDMaE}
new_design_DDMaE <- design(data = dat,
        contrasts = list(S = Smat),
        formula=list(v~S,a~E, t0~1, s~1, Z~1, sv~1),
        constants=c(s=log(1)),
        model=DDM)
```

Let's examine the parameter mapping:

```{r mapped_pars_new_design_DDMaE}
mapped_pars(new_design_DDMaE)
```

The intercept (v) represents the drift bias - the average of rates for
left and right stimuli:

-   Negative values indicate bias towards left-moving stimuli (lower
    boundary)

-   Positive values indicate bias towards right-moving stimuli (upper
    boundary)

-   Zero represents no bias

The difference parameter (v_Sdif) represents the conventional "drift
rate":

-   Positive values indicate stronger accumulation towards correct
    responses

-   Higher values suggest better discrimination ability between stimuli

### Design Specification

Now we can use this contrast in a model where drift rate (v_Sdif) varies
with E. We expect that accuracy emphasis causes participants to pay more
attention to the stimuli and better discriminate left and right stimuli.

Since response emphasis is unlikely to affect rate bias, we'll fix bias
effects for speed vs. neutral (v_Eneutral) and speed vs. accuracy
(v_Eaccuracy) to zero in the constants argument. This means v will index
the same rate bias across all E levels. Since both a and v are
influenced by E, we call this the DDMavE design:

```{r design_DDMavE}
design_DDMavE <- design(data = dat,model=DDM,
                        contrasts = list(S = Smat),
                        formula=list(v~S*E,a~E, t0~1, s~1, Z~1, sv~1),
                        constants=c(s=log(1), v_Eneutral = 0, v_Eaccuracy = 0))
```

Let's have a look at how our parameters map to the cells of the data:

```{r mapped_pars_design_DDMavE}
mapped_pars(design_DDMavE)
```

### Prior Specification

For the hierarchical model we specify:

1.  A prior on the group-level mean
2.  A prior on the variance-covariance matrix

For the group-level mean, we need to specify prior means and standard
deviations for each parameter. We can reuse values from the previous
prior for the single subject model by supplying the previous prior to
the update argument. This means we only need to specify values for the 4
new parameters.

For the intercept and drift rates, we'll use values that yield the same
implied effect on left and right rates as in the previous model. We set
`type = standard` to indicate this is a prior for a hierarchical model
with full variance-covariance matrix:

```{r prior_DDMavE, fig.width = 7, fig.height = 5}
prior_DDMavE <- prior(design_DDMavE, update = prior_DDMaE,
                    mu_mean=c(v=0,v_Sdif=2,'v_Sdif:Eneutral'=0.1,
                              'v_Sdif:Eaccuracy'=0.2),
                    mu_sd=c(v=1,v_Sdif=1,'v_Sdif:Eneutral'=1,
                            'v_Sdif:Eaccuracy'=1), type = "standard")

# Let's check our choices look reasonable
plot(prior_DDMavE)
```

We can also give the prior to the plot_design function. This gives
analogous results to plot(design), but uses the prior mean as the
parameter input:

```{r plot_design_DDMavE, fig.height=6, fig.width = 7}
plot_design(prior_DDMavE, factors = list(v = c("S", "E"), a = "E"), 
            plot_factor = "S")
```

We can also specify priors on the group-level covariance matrix. This
prior is best considered in two parts:

1.  Variances: Default priors are vague but favor smaller values
2.  Correlations: Default priors are uniform on the -1 to 1 interval

Note that here the default is mapped=FALSE and there is no mapped=TRUE
option, since the variances cannot be easily mapped to the natural
scale.

We only plot the prior on the correlations between other parameters and
the `v` parameter to reduce the output, but the prior is the same across
all correlations.

```{r plot_prior_DDMavE_sigma2, fig.width = 7, fig.height = 5}
plot(prior_DDMavE, selection = "sigma2", layout = c(3,3))
plot(prior_DDMavE, selection = "correlation", use_par = "v")
```

The variance-covariance prior settings can be adjusted using two
hyperparameters: - Parameter A controls the variance width - Parameter v
controls the correlations

For example, we can make the correlation prior more peaked at zero by
increasing v from its default of 2 to 4. However, this flexibility is
limited as the same prior applies to all correlations. We can however, set 
different priors for the variances using the scale parameter A (default 0.3). 

```{r plot_prior_DDMavE_correlation, fig.width = 7, fig.height = 5}
new_prior <- prior(design_DDMavE,v=4, A = c("a" = .5), update=prior_DDMavE)
plot(new_prior, selection = "correlation", use_par = "v")
plot(new_prior, selection = "sigma2", use_par = c("v", "a"))
```

To get more details on the prior structure you can use the summary
function, which will now include information on the prior on the
group-level mean and group-level variance:

```{r}
summary(prior_DDMavE)
```

We can also plot the "implied" priors at the individual level by
sampling from the group-level distribution. This prior depends on the
group level distribution, so when the group level distribution is
updated during sampling, the implied prior becomes less vague. This
creates the "shrinkage" effect that makes hierarchical modeling
valuable:

```{r plot_prior_DDMavE_alpha, fig.width = 7, fig.height = 5}
plot(prior_DDMavE, selection = "alpha", layout = c(2,3))
```

### Model Estimation

We now fit the model using two levels of parallelization:

1.  Across chains: Default behavior (3 chains)
2.  Within chain: Set cores_per_chain = 4

This results in 12 total cores (3 chains × 4 cores per chain). Note that
within-chain parallelization is not available on Windows systems.

```{r make_emc_DDMavE, eval=FALSE}
DDMavE <- make_emc(dat, design = design_DDMavE, prior = prior_DDMavE)
DDMavE <- fit(DDMavE, cores_per_chain = 4, fileName = "tmp.RData")
save(DDMavE, file = "DDMavE.RData")
```

```{r load_DDMavE, eval=TRUE, echo=FALSE}
load("DDMavE.RData")
```

After sampling we again check for convergence.

```{r check_DDMavE}
check(DDMavE)
```

### Model Inference

#### Model Fit

Let's first inspect posterior predictives. We set n_cores = 12 to speed
up computations:

```{r predict_DDMavE}
pp_DDMavE <- predict(DDMavE, n_cores = 12)
```

The model seems to fit reasonably well, but there's still an
underestimation of the proportion of error responses, most notably in
the neutral condition.

```{r plot_cdf_DDMavE}
plot_cdf(DDMavE, pp_DDMavE, factors = "E", functions = c(correct = acc_fun), 
         layout = c(1,3),defective_factor = "correct", 
         legendpos = c("topright", "topleft"))
```

Comparing this model to the single subject data shows improved fit
across subjects. The quantiles of the posterior predictive distribution
are much tighter (barely visible on the plot), which is expected given
the larger amount of data being fit.

We can also directly compare the fit of the new model to the old model
for only the first subject (since the first model is only fit to the
first subject):

```{r plot_cdf_DDMavE_subject}
plot_cdf(DDMaE, list(a = pp_DDMaE, 'a&v' = pp_DDMavE), factors = "E", 
         functions = c(correct = acc_fun), layout = c(1,3),
         defective_factor = "correct", 
         legendpos = c("topright", "topleft"), subject = 1)
```

Note that despite the second model being more flexible, it is possible
it provides a poorer fit to the data for this single subject due to
shrinkage effects.

#### Model Comparison

Now that we've examined model fit, we can test research questions.

For example, is there support for (conventional) threshold and (less
conventional) drift rate differences across emphasis conditions at the
population (mu) level?

We can plot the implied accumulation process of our estimated parameters
using the `plot_design` function, but now with the emc object:

```{r design_plot_samples_DDMavE}
plot_design(DDMavE, factors = list(v = c("E", "S"), a= "E"), plot_factor = "S")
```

This suggests small differences in drift rates and thresholds between
the neutral and accuracy conditions.

To get a better understanding of the effect, let's plot the posterior
parameter densities for our parameters of interest:

```{r plot_pars_DDMavE_mu}
plot_pars(DDMavE, selection = "mu", 
          use_par = c("a_Eneutral", "a_Eaccuracy", 
                      "v_Sdif:Eneutral", "v_Sdif:Eaccuracy"))
```

The following 95% credible intervals clearly show strong effects on both
rates and thresholds, with similar magnitudes for neutral and accuracy
effects:

```{r plot_credint_DDMavE_mu}
credint(DDMavE, use_par = c("a_Eneutral", "a_Eaccuracy", 
                            "v_Sdif:Eneutral", "v_Sdif:Eaccuracy"))
```

We can also visualize the individual differences in these effects by
setting `all_subjects = TRUE`:

```{r plot_pars_DDMavE_all}
plot_pars(DDMavE, layout = c(2,2),
          use_par = c("a_Eneutral", "a_Eaccuracy",
                      "v_Sdif:Eneutral","v_Sdif:Eaccuracy"),
          all_subjects = TRUE)
```

Here the group-level distribution (simulated from the estimated
group-level mean and variance) is in red and the individual-level
posteriors are in black.

Essentially the group-level distribution acts as the prior to the
individual-level parameters. Notice that threshold effects are almost
all above 0, whereas drift rate effects sometimes overlap zero,
indicating that some participants do not pay more attention with greater
accuracy emphasis.

We can compute Bayes Factors for these parameter differences using the
savage-dickey ratio, testing the group-level mean differences in drift
rates and thresholds between the Neutral and Accuracy conditions. Values
less than one favor the null hypothesis, and values greater than one
favor the alternative hypothesis.

```{r vdiff_DDMavE, fig.width = 7, fig.height = 4}
vdiff <- function(p)diff(p[c("v_Sdif:Eneutral","v_Sdif:Eaccuracy")])
adiff <- function(p)diff(p[c("a_Eneutral","a_Eaccuracy")])
par(mfrow=c(1,2))
hypothesis(DDMavE,fun=vdiff)
hypothesis(DDMavE,fun=adiff)
```

The plot shows the posterior (black) and prior (red) densities. The
circles are plotted at the height of the densities at H0 (in this case
0), since we're testing whether the differences between these parameters
are larger than 0. If the prior is higher than the posterior at H0 that
means there's evidence for the effect, if lower evidence against it.

We find strong evidence against the drift rate difference between the
neutral and accuracy condition, and more moderate evidence against the
threshold difference. Thus suggesting that these conditions have similar
effects on the drift rate and threshold settings. Similar conclusions
are reached based on 95% credible intervals for these differences.

```{r credible_DDMavE_mu}
credible(DDMavE, x_fun=vdiff)
credible(DDMavE,x_fun=adiff)
```

# Linear Ballistic Accumulator Model (LBA)

### Design Specification

We will now try to incorporate our previous findings in an LBA model.
While we would normally do full model development for the DDM and LBA
separately, this tutorial is just for illustrative purposes.

As the LBA is a race model it has one accumulator representing each
possible response, in this case separate accumulators for the left and
right response. EMC2 uses the levels of the "R" factor found in the data
to construct a factor which denotes the accumulators "lR" (the latent
response) with level names taken from Rlevels.

The lR factor allows for response bias, analogous to Z in the DDM. In
race models, response bias can be modeled by allowing different
thresholds (B) for different responses. Here we will use B\~lR to allow
for different thresholds for the accumulator corresponding to left and
right stimuli (e.g., a bias to respond left occurs if the left threshold
is lower than the right threshold).

For race models, you often need to provide a "matchfun" function that
marks which accumulator matches the correct response. Typically, this
function takes the latent response (lR) and returns TRUE when it matches
the stimulus (S):

```{r matchfun_LBA}
matchfun=function(data)data$S==data$lR
```

matchfun is used to automatically to create a latent match (lM) factor
with levels "FALSE" (i.e., the stimulus does not match the accumulator)
and "TRUE" (i.e., the stimulus does match the accumulator). This is
added internally and can also be used in model formula, typically for
parameters related to the rate of accumulation.

When working with lM, it's often useful to design an "average and
difference" contrast matrix. For binary responses, this has a simple
form:

```{r ADmat_LBA}
ADmat <- matrix(c(-1/2,1/2),ncol=1,dimnames=list(NULL,"d"))
ADmat
```

When applied to drift rates (v\~lM), the contrast parameter v_lMd
represents the difference between correct and incorrect response drift
rates. This "rate quality" parameter is analogous to the DDM drift rate.
When accuracy exceeds chance, rate quality is positive.

The ADmat intercept (v) represents the average accumulation rate across
accumulators, measuring both stimulus magnitude and decision urgency
effects. Unlike the DDM, which represents only evidence differences
between possible responses, this "rate quantity" parameter captures the
overall tendency to respond. The DDM does not directly represent this
tendency.

For simplicity, we omit rate bias, although it could be included by
allowing v\~S\*lM.

We also allow the drift rate standard deviation to vary as a function of
match (sv\~lM), as has become standard in more recent applications of
the LBA.

Just as with the DDM, we fix one parameter to make the model
identifiable; in this case, as is common, we chose to fix drift rate
standard deviation intercept (sv).

Given the findings of the hierarchical DDM that there's no evidence for
a threshold or drift rate difference between neutral and accuracy
emphasis, we construct our model to estimate one shared E parameter for
the neutral and accuracy condition. To achieve this, we use the
functions argument.

```{r E2_LBA}
E2 <- function(data) factor(data$E!="speed",labels=c("speed","nonspeed"))
design_LBABvE <- design(data = dat,model=LBA,matchfun=matchfun,
                       formula=list(v~lM*E2,sv~lM,B~E2+lR,A~1,t0~1),
                       contrasts=list(lM = ADmat),constants=c(sv=log(1)),
                       functions=list(E2=E2))
```

Here the E2 function creates a new factor in the data, with levels speed
and nonspeed.

Whereas the drift rate intercept in the DDM represents response bias,
which is unlikely to be affected by emphasis instructions, in the LBA
the intercept codes for urgency effects, which are more likely to be
affected by emphasis instructions. We therefore allow the intercept to
also vary by E, yielding a V_E2nonspeed parameter and a v_LMd:E2nonspeed
parameter. The latter parameter represents the difference in
accumulation rate between correct and incorrect responses under
neutral/accuracy relative to the speed emphasis condition.

Let's examine how our parameters map to the design:

```{r mapped_pars_design_LBABvE}
mapped_pars(design_LBABvE)
```

### Prior Specification

We now create a prior. Note that all parameters except v are estimated
on a log scale (see ?LBA). For the LBA, t0 is typically smaller than for
the DDM, so we set it to 0.2. We further set increasing thresholds and
urgency for higher response caution conditions. As we have no response
bias expectation, B_lRright is set to zero.

```{r prior_LBABvE, fig.width = 7, fig.height = 5}
mu_mean <- c(v=1, v_E2nonspeed = -.2, v_lMd=1, "v_lMd:E2nonspeed"=.2,
             sv_lMd=log(1),B=log(1), B_E2nonspeed = log(1.5), B_lRright=0,  
             A=log(0.25), t0=log(.2))
mu_sd <- c(v=1, v_lMd=0.5, "v_lMd:E2nonspeed"=0.5, 
           sv_lMd=.5,B=0.3, B_E2nonspeed=0.3,B_lRright=0.3, A=0.4, t0=.5)

prior_LBABvE <- prior(design_LBABvE, type = 'standard',mu_mean=mu_mean,
                      mu_sd=mu_sd)
plot(prior_LBABvE, layout = c(2,3))

plot_design(prior_LBABvE, factors = list(v = c("E2", "lM"), B = "E2"), 
            plot_factor = "E2")
```

### Model Estimation

Now we can fit the model:

```{r make_emc_LBABvE, eval = FALSE}
LBABvE <- make_emc(dat,design_LBABvE, prior=prior_LBABvE)
LBABvE <- fit(LBABvE, cores_per_chain = 4, fileName="tmp.RData")
save(LBABvE,file="LBABvE.RData")
```

```{r load_LBABvE, eval=TRUE, echo=FALSE}
load("LBABvE.RData")
```

The model shows good convergence and generally good efficiency:

```{r check_LBABvE}
check(LBABvE)
```

### Model Inference

#### Model Fit

We generate posterior predictions for the LBA model to see how well the
model fits the data:

```{r predict_LBABvE}
pp_LBABvE <- predict(LBABvE, n_cores = 12)
plot_cdf(LBABvE, pp_LBABvE, factors = "E2", functions = c(correct = acc_fun),
         layout= c(1,2), defective_factor = "correct", 
         legendpos = c("right","topleft"))
```

We can also directly compare the fit of the LBA model to the DDM model:

```{r plot_cdf_LBABvE_DDMavE, echo = FALSE}
plot_cdf(LBABvE, list(LBA = pp_LBABvE, DDM = pp_DDMavE), factors = "E", 
         functions = c(correct = acc_fun), layout = c(1,3),
         defective_factor ="correct", legendpos = c("right", "topleft"))
```

The models show different patterns of misfit. The DDM underestimates the
number of error responses, especially in the neutral condition. In
contrast, the LBA model slightly overestimates error responses in the
accuracy condition. Note that since we specified that the LBA model
should only estimate one E parameter for the neutral and accuracy
condition, the LBA posterior predictives do not show a difference
between these conditions. Although the DDM includes that flexibility, it
does not include separate urgency effects for the neutral and accuracy
condition (as it does not include urgency effects).

We can also compare arbitrary descriptives on the real data to the
predictives of both models. For example differences in response time and
error rates between the emphasis conditions:

```{r plot_stat_LBABvE, fig.width= 8, fig.height = 5}
drt_ns <- function(data){
  all <- tapply(data$rt,data$E,mean)
  out <- all['neutral'] - all['speed']
  names(out) <- "Neutral - Speed"
  return(out)
}
drt_as <- function(data){
  all <- tapply(data$rt,data$E,mean)
  out <- all['accuracy'] - all['speed']
  names(out) <- "Accuracy - Speed"
  return(out)
}
derr_ns <- function(data){
  data$correct <- data$S == data$R
  all <- tapply(data$correct,data$E,mean)*100
  out <- all['neutral'] - all['speed']
  names(out) <- "Neutral - Speed"
  return(out)
}
derr_as <- function(data){
  data$correct <- data$S == data$R
  all <- tapply(data$correct,data$E,mean)*100
  out <- all['accuracy'] - all['speed']
  names(out) <- "Accuracy - Speed"
  return(out)
}
par(mfrow = c(2,2))
plot_stat(LBABvE, list(LBA = pp_LBABvE, DDM = pp_DDMavE), stat_fun = drt_ns, 
          legendpos = c('topleft', 'topright'), main = "Neutral - Speed",
          layout=NULL,xlim=c(.08,.16),adjust=1.5,xlab = "RT (sec)")
plot_stat(LBABvE, list(LBA = pp_LBABvE, DDM = pp_DDMavE), stat_fun = drt_as, 
          legendpos = c('topleft', 'topright'), main = "Accuracy - Speed",
          layout=NULL,xlim=c(.08,.16),adjust=1.5,xlab = "RT (sec)")
plot_stat(LBABvE, list(LBA = pp_LBABvE, DDM = pp_DDMavE), stat_fun = derr_ns, 
          legendpos = c('topleft', 'topright'), main = "Neutral - Speed",
          layout=NULL,xlim=c(7,16),adjust=1.5, xlab = "Accuracy (%)")
plot_stat(LBABvE, list(LBA = pp_LBABvE, DDM = pp_DDMavE), stat_fun = derr_as, 
          legendpos = c('topleft', 'topright'), main = "Accuracy - Speed",
          layout=NULL,xlim=c(7,16),adjust=1.5, xlab="Accuracy (%)")
```


Here it is clear that the LBA model fits better in terms of accuracy
differences, whereas both models have similar quality accounts of RT
differences, but with opposite misfit for the accuracy - speed
difference.

#### Model Comparison

Although visual fit is helpful in assessing absolute model fit,
statistical differences between the models can be better tested using
Bayes Factors. These take into account how well the model fits all
aspects of the entire data set, while penalizing for model complexity.

EMC2 estimates Bayes Factors between sets of samples using the Warp-III
bridge sampling method. To assess numerical stability see `both_splits` in
`run_bridge_sampling`.

```{r compare_LBABvE_DDMavE}
compare(list(LBA = LBABvE, DDM = DDMavE), cores_for_props = 3)
```

Here the MD (Marginal Deviance; used for Bayes Factors), DIC (Deviance
Information Criterion) and BPIC (Bayesian Predictive Information
Criterion) are measures of model fit. The lower the better. The weights
(wX) are the probabilities of the models, the higher the better.

The model comparison indicated that the LBA model provides a better
account of the data. We can now use our winning model to do some further
hypothesis testing. For example, how much evidence do we have for
threshold, quality and quantity differences between the speed and
non-speed conditions?

```{r diff_hypothesis_LBABvE, fig.width = 7, fig.height = 4}
b_diff <- function(p)diff(p[c("B","B_E2nonspeed")])
qual_diff <- function(p)diff(p[c("v_lMd","v_lMd:E2nonspeed")])
quant_diff <- function(p)diff(p[c("v","v_E2nonspeed")])
par(mfrow=c(1,3))
hypothesis(LBABvE,fun=b_diff)
hypothesis(LBABvE,fun=qual_diff)
hypothesis(LBABvE,fun=quant_diff)
```

The results show very strong evidence for threshold and urgency
differences but only moderate evidence for a quality difference.

To see the direction of these differences:

```{r credint_LBABvE, fig.width = 7, fig.height = 4}
credint(LBABvE, use_par = c("v_E2nonspeed", "v_lMd:E2nonspeed", "B_E2nonspeed"))
```

Lastly, we can test for evidence of response bias:

```{r bias_LBABvE, fig.width = 7, fig.height = 4}
hypothesis(LBABvE,"B_lRright")
```

Evidence for bias is equivocal bias, with a slight tendency favoring the
left response (i.e., a higher threshold for right).

We finish by inspecting the individual differences among the parameters
of interest:

```{r plot_pars_LBABvE_all}
plot_pars(LBABvE, all_subjects = TRUE, use_par = c("B_lRright", "B_E2nonspeed",
                                            "v_E2nonspeed", "v_lMd:E2nonspeed"))
```

Everybody seems to have an E threshold effect, quality and quantity
effects are more variable, and biases even more so.

This tutorial is not intended as a complete model comparison. The DDM
could have potentially been more competitive in terms of complexity
penalized model selection with similar parameter constraints on the
emphasis effects to the LBA, and improved in absolute fit by allowing
for additional between-trial variability in the parameters (st0 and SZ).
Similarly, for the LBA, the misfit to RT differences between the speed
condition and the other two conditions could be addressed by allowing
rates and/or thresholds to differ between the accuracy and neutral
conditions.
